Think deep for reef resilience

### 2025-08-11 paper/paper_draft.qmd
Manually update GBRMPA citations

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change
[@bozec2022]. These coral cover declines will be compounded by ocean acidification,
intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012;
@fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass
bleaching event may be signalling a shift towards this reef future, with some level of
bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@cantin2024].

@@

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. An indicative map of the GBR marine park including the Marine Park management
areas [@GBRMPA2024] and the corresponding bioregion for each reef is shown in
[@fig-region-map; @GBRMPA2001]. This investigation adopts the GBR reef outlines as published
by GBRMPA [-@GBRMPA2022], which defines 3806 reefs across the GBR.

@@

While data adopted for the given assessment are said to be accurate up to 15m from Mean Sea
Level [@GBRMPA2020], there are many sources of uncertainty in the depth estimates used for
this study. These are outlined further in the Discussion. Notably, the adopted depth
estimates were derived from bathymetries based on satellite imagery. Available data can be
limited in their alignment with the established GBRMPA reef outlines used to characterise
reef locations and their areal extent. Quality control flags were assigned to identify
potential data limitations: reefs with no overlapping bathymetry data, reefs with minimum
depth values above sea level (indicating potential data artifacts), and reefs where
bathymetry pixels covered less than 5% of the reef polygon area. This removed 842 reefs, or
22%, from the initial set of 3806 reefs. Furthermore, reefs that are not associated with
GBRMPA management areas [@GBRMPA2024] or reefal bioregions [@GBRMPA2001] were also removed
from the investigation. Bioregions which contained fewer than three reefs were also
excluded. As a result, 1377 reefs were excluded (36%), leaving a total of 2429 remaining for
the purpose of this investigation.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points). Colour-coded areas outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2024]. Reef centroids are marked individually for visual clarity, and are coloured
according to its bioregion [@GBRMPA2001]. All 2429 reefs included in this investigation are
represented in this figure. An annotation represents the approximate location of the
Australian Institute of Marine Science Townsville site at Cape
Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map
fig-align="center" width="14.8cm"}

@@

Timeseries clustering analyses were performed at the management area and reefal bioregion
scale (defined in GBRMPA [-@GBRMPA2001]). Additional analyses were also performed at the
whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown
here. To compare timeseries clustering results across GCMs, management area scale clustering
results are used to preserve coarse scale latitudinal differences between regions, while
maintaining ease of comparison and visualisation. These comparisons include quantification
of differences between low and high cover clusters across management regions under each GCM
using the Complexity-Invariant Distance between cluster median timeseries. To assess how
different reef characteristics influence timeseries clustering at a finer scale, bioregion
results are used to capture patterns occurring within individual clusters of reefs. Results
of all analyses at all clustering scales can be found at \[**insert citation for
repository**\] for those interested.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton], for example, identified 120 false
positives, where features identified as reefs were in actuality sand banks or rocky reefs
(rather than coral reefs). Reefs with bathymetry data of less than 5% of its indicative area
were excluded from analysis to ensure analyses were not conducted on reefs with no data
coverage (see @sfig-depth-cutoff). The existing GBRMPA reef outlines were adopted to align
with other ongoing modelling and analyses conducted as part of the Reef Restoration and
Adaptation Program (RRAP; <https://gbrrestoration.org>) **\[TODO: PROPER CITATION OF
RRAP\]**. The median depths for each indicative reef area were extracted from bathymetry
data using zonal statistics and depth values beyond the 15m range were included. The
resulting depth estimates are therefore highly uncertain and are only adopted as a useful
indication for the purpose of the presented analysis. They are not intended to be an
accurate and precise report of reef depths across the GBR.

@@

Further analyses could build on the findings of this study by analysing the impacts of
prioritising reef depth in restoration intervention decision making. The ADRIA platform is
developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and
ecosystem scenario modelling into reef restoration decision support. The platform allows for
the investigation of different reef characteristics to be targetted in decision making and
models the evolving outcomes as a result of interventions at various locations. The
capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of
waves, cyclones, CoTS and currents into future resilience investigations [@lin2023;
@mcclanahan2005; @grimaldi2022; @liu2024]. Additionally, the investigation could be enhanced
by incorporating ocean acidification projections into analyses of coral carbonate budgets.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by management area scale cluster assignment.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2024]. Reef centroids are
marked individually for visual clarity. Reefs that do not have a consistent cluster
assignment level across all GCMs at mangement area scale are gray. All consistently
clustered reefs are coloured according to their management area scale cluster assignment.
Green represents reefs in the low cover cluster, orange represents reefs in the medium cover
cluster and blue represents reefs in the high cover cluster. All 2429 reefs included in this
investigation are represented in this figure. An annotation represents the approximate
location of the Australian Institute of Marine Science Townsville site at Cape
Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/gcm_cluster_assignment_map.png){#sfig-gcm-cluster-map
fig-align="center" width="14.82cm"}

>>>>

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change
[@bozec2025; @bozec2022]. These coral cover declines will be compounded by ocean
acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs)
[@cantin2024].

@@

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. An indicative map of the GBR marine park including the Marine Park management
areas [@GBRMPA2007] and the corresponding bioregion for each reef is shown in
(@fig-region-map; [@GBRMPA2024]). This investigation adopts the GBR reef outlines as published
by GBRMPA [-@GBRMPA2022], which defines 3806 reefs across the GBR.

@@

While data adopted for the given assessment are said to be accurate up to 15m from Mean Sea
Level [@GBRMPA2020], there are many sources of uncertainty in the depth estimates used for
this study. These are outlined further in the Discussion. Notably, the adopted depth
estimates were derived from bathymetries based on satellite imagery. Available data can be
limited in their alignment with the established GBRMPA reef outlines used to characterise
reef locations and their areal extent. Quality control flags were assigned to identify
potential data limitations: reefs with no overlapping bathymetry data, reefs with minimum
depth values above sea level (indicating potential data artifacts), and reefs where
bathymetry pixels covered less than 5% of the reef polygon area. This removed 842 reefs, or
22%, from the initial set of 3806 reefs. Furthermore, reefs that are not associated with
GBRMPA management areas [@GBRMPA2007] or reefal bioregions [@GBRMPA2024] were also removed
from the investigation. Bioregions which contained fewer than three reefs were also
excluded. As a result, 1377 reefs were excluded (36%), leaving a total of 2429 remaining for
the purpose of this investigation.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points). Colour-coded areas outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity, and are coloured
according to its bioregion [@GBRMPA2024]. All 2429 reefs included in this investigation are
represented in this figure. An annotation represents the approximate location of the
Australian Institute of Marine Science Townsville site at Cape
Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map
fig-align="center" width="14.8cm"}

@@

Timeseries clustering analyses were performed at the management area and reefal bioregion
scale (defined in GBRMPA [-@GBRMPA2024]). Additional analyses were also performed at the
whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown
here. To compare timeseries clustering results across GCMs, management area scale clustering
results are used to preserve coarse scale latitudinal differences between regions, while
maintaining ease of comparison and visualisation. These comparisons include quantification
of differences between low and high cover clusters across management regions under each GCM
using the Complexity-Invariant Distance between cluster median timeseries. To assess how
different reef characteristics influence timeseries clustering at a finer scale, bioregion
results are used to capture patterns occurring within individual clusters of reefs. Results
of all analyses at all clustering scales can be found at \[**insert citation for
repository**\] for those interested.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton], for example, identified 120 false
positives, where features identified as reefs were in actuality sand banks or rocky reefs
(rather than coral reefs). Reefs with bathymetry data of less than 5% of its indicative area
were excluded from analysis to ensure analyses were not conducted on reefs with no data
coverage (see @sfig-depth-cutoff). The existing GBRMPA reef outlines were adopted to align
with other ongoing modelling and analyses conducted as part of the Reef Restoration and
Adaptation Program \[RRAP; <https://gbrrestoration.org>\] **\[TODO: PROPER CITATION OF
RRAP\]**. The median depths for each indicative reef area were extracted from bathymetry
data using zonal statistics and depth values beyond the 15m range were included. The
resulting depth estimates are therefore highly uncertain and are only adopted as a useful
indication for the purpose of the presented analysis. They are not intended to be an
accurate and precise report of reef depths across the GBR.

@@

Further analyses could build on the findings of this study by analysing the impacts of
prioritising reef depth in restoration intervention decision making. The ADRIA platform is
developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and
ecosystem scenario modelling into reef restoration decision support. The platform allows for
the investigation of different reef characteristics to be targetted in decision making and
models the evolving outcomes as a result of interventions at various locations. The
capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of
waves, cyclones, CoTS and currents into future resilience investigations [@lin2023;
@mcclanahan2005; @grimaldi2023; @grimaldi2022; @liu2024]. Additionally, the investigation
could be enhanced by incorporating ocean acidification projections into analyses of coral
carbonate budgets.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by management area scale cluster assignment.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Reefs that do not have a consistent cluster
assignment level across all GCMs at mangement area scale are gray. All consistently
clustered reefs are coloured according to their management area scale cluster assignment.
Green represents reefs in the low cover cluster, orange represents reefs in the medium cover
cluster and blue represents reefs in the high cover cluster. All 2429 reefs included in this
investigation are represented in this figure. An annotation represents the approximate
location of the Australian Institute of Marine Science Townsville site at Cape
Cleveland.](../outputs/ADRIA_results/HighResCoralStress/figs/gcm_cluster_assignment_map.png){#sfig-gcm-cluster-map
fig-align="center" width="14.82cm"}
